---
title: 'dplr: Practice'
author: "Martha Miller"
output:
  html_document:
    css: ../workshop.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(warning=FALSE, message=FALSE)
```


### Earthquake data

```{r load-quakes, eval = FALSE}
data(quakes)
```

1.   Show a command you could use to find the average (mean) magnitude of all earthquakes that occurred below the median depth of all earthquakes.

```{r mean magnitude of deep earthquakes}
library(ggplot2)

deep <- quakes %>%
  select(depth, mag) %>%
  filter(depth < median(depth))%>%
  summarize(avg_mag = mean(mag, na.rm = TRUE)) 
deep
```


1.   How does the average magnitude of these "deep" earthquakes compare to the magnitude of "shallow" (above median depth) earthquakes?
```{r shallow earthquakes}
shallow <- quakes %>%
  select(depth, mag) %>%
  filter(depth > median(depth))%>%
  summarize(avg_mag = mean(mag, na.rm = TRUE)) 
shallow
```

1.   Show the command you would use to get the row(s) with the smallest value in the `depth` column.
```{r smallest depth}
quakes%>%
  filter(depth == min(depth))
```

1.   Show the command you would use to create a new data frame named `quakes_sorted` that stores the data in sorted order by `depth`.


```{r dataframe of sorted quakes}
quakes_sorted <- quakes %>%
  arrange(depth)

str(quakes_sorted)
```


* * *

### Flights data

```{r load-flights, eval = FALSE}
library(pnwflights14)
data(flights)
```

1. Say I fly very often between PDX airport in Portland and Logan Airport in Boston. Based on the 2013 data, which day of the week would you recommend that I buy a ticket for if I want to minimize delays?
```{r weekday flight}
str(flights)
summary(flights$day)

f <- flights %>%
  filter(origin == "PDX" & dest == "BOS" | origin == "BOS" & dest == "PDX") %>%
  mutate(date = as.Date(paste(day, "-", month, "-", year, sep = ""), '%d-%m-%Y')) %>%
  mutate(weekdays_name = weekdays(date))
str(f)
names(flights)

f %>%
  group_by(weekdays_name) %>%
  summarize(avg_delay = mean(dep_delay + arr_delay, na.rm=TRUE)) %>%
  arrange(avg_delay)
  
summary(f$dep_delay)
```

Based on average total delay between arrival and departure times, Saturday is the best day to fly, followed by Sunday and Monday. These three days of the week have negative average delays, indicating that they are more likely to leave and arrive ahead of schedule.

1. What advice do you have for the best airline for this route?

```{r best airline}
airlines <- flights %>%
  filter(origin == "PDX" & dest == "BOS" | origin == "BOS" & dest == "PDX") %>%
  group_by(carrier) %>%
  summarize(avg_delay = mean(dep_delay + arr_delay, na.rm=TRUE)) %>%
  arrange(avg_delay)

airlines 
```
We recommend flying on the airline Alaska, as the average total delay when flying between Portland and Boston is negative, indicating it tends to be ahead of schedule more than other airlines on this route. There actually appears to be a choice between only Alaska and JetBlue.
* * *

### Back to Global Warming

Go back to the vector `tData` that you used for the project. Use this data and
maybe `dplyr` to form a dataframe that will make the analysis much more straightforward.
```{r global warming}
library(dplyr)
load(url('http://www.stolaf.edu/people/olaf/cs125/project1.RData'))
temp <- data.frame(tData)
absolute_farhenheit <- function(scaled_celsius){
  descaled_celsius <- scaled_celsius / 100
  celsius <- descaled_celsius + 14
  farhenheit <- celsius *1.8 + 32
  return (farhenheit)
}
temp <- absolute_farhenheit(temp)

months <- (as.ordered("January", "February", "March", "April", "May", "June", "July", "August", "September","October", "November", "December"))
year <- rep(1881:2012, each = 12)

df <- data.frame(temp, months, year)
```
