---
title: "Stringr"
author: Em Beauchamp, Some material adapated from stringr vignette
output: 
  ioslides_presentation:
      incremental: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stringr)
library(dplyr)
```
## Background Information {.build}

- Written to fill in the gap between other languages' (i.e. Python) string capability and R. 
- Wrapper to Stringi
- Acts as an interface to other languages
- Consistent Syntax
- 


## Usefulness of String Manipulations in R {.build}
```{r, echo = TRUE}
human_1 <-"A human would recognize this as the same sentence"
human_2 <- "A HUMAN WOULD RECOGNIZE THIS AS THE SAME SENTENCE"


human_1 == human_2

str_to_lower(human_1) == str_to_lower(human_2)
```

## Basic String Operations {.build}
```{r, echo = TRUE}
some_text_vector <- c("Red", "BLUE", "orange", "Purple")

some_text_vector

some_text_vector <- str_to_title(some_text_vector)

some_text_vector
```
Also: str_pad, str_join, str_trim, etc.

## String Replacement Operations {.build}

- Stringr also handles string matching and regular expression matching
- Perhaps a useful tool : http://regexr.com/
```{r echo = TRUE}
text_input_vector <- c(">Some", ">Words")
text_input_vector
```

```{r echo = TRUE}

text_input_vector_clean <- str_replace(text_input_vector, ">", "")
text_input_vector_clean

```

## Plays Well With The Pipe!! {.build}
```{r echo = TRUE}
text_input_vector %>%
  str_replace(">", "") %>%
  str_to_lower()

```

## More Complex Example {.build}
https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html







