---
title: "Dataframes: Practice"
author: "Martha Miller"
output:
  html_document:
    css: ../workshop.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(warning = FALSE, message = FALSE)
```

For the following exercises, load in the built in earthquake data set.

```{r, eval=FALSE}
data(quakes)
```

1.   How would you find information about this data frame? How many rows and columns does it have?
```{r}
dim(quakes)
str(quakes)
```

1.   Based purely on the description of the columns (found with `?quakes`), would you expect a relationship between the 4th and 5th columns? Why or why not?

Columns "mag" and "stations" provide information about the Richter Magnitude and Number of stations reporting. I don't know much about earthquakes but I would expect the potential relationship to reflect that maybe there are more stations near earthquake 'hotspots;' in other words, I think there might be a positive relationship.

1.   Show a command you could use to find the average (mean) magnitude of all earthquakes that occurred below the median depth of all earthquakes.

```{r}
mean(quakes$mag[quakes$depth < median(quakes$depth)], na.rm =TRUE)
```


1.   How does the average magnitude of these "deep" earthquakes compare to the magnitude of "shallow" (above median depth) earthquakes?

```{r}
mean(quakes$mag[quakes$depth > median(quakes$depth)], na.rm =TRUE)
```
Deep earthquakes, on average, have a larger magnitude than shallow earthquakes.


1.   Show the command you would use to get the row(s) with the smallest value in the `depth` column.

```{r}
which.min(quakes$depth)

#check stats for column
summary(quakes$depth)
```



<div id="license">
Adapted from material developed by Olaf Hall-Holt and Paul Roback for CS 125 at St. Olaf College.
</div>