---
title: "Vectors: practice II"
author: "Martha Miller"
output:
  html_document:
    css: ../workshop.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(warning=FALSE,message=FALSE)
```


## Sorting Vectors

1.   

In addition to its use on vectors, `order()` can also be used to sort (or
I suppose order!) dataframes.  Consider one of R's preloaded datasets, from
Motor Trend Cars.

```{r}
data(mtcars)
mtcars
```

say you wanted to look through this list of cars and order them in a way in a way
that showcases which ones have the best `mpg`, then, the ones that have the 
highest `hp`.  Using `order()`, rearrange this data frame such that it's ordered
first by `mpg` (low to high) and then by `hp` (high to low).
```{r}
mpg.mtcars <- mtcars[order(mtcars$mpg), ]
mpg.mtcars
hp.mtcars<- mtcars[order(mtcars$hp, decreasing = TRUE),]
hp.mtcars
```

* * *

## Changing Vector Elements

Please paste the following commands into your Rstudio session, in order to set up the following exercises.

```{r}
require('ggplot2')
load(url('http://www.stolaf.edu/people/olaf/cs125/hw9.RData'))
```

Let's see whether the drawing function is working correctly. Try pasting in the following command, and see what it does:

```{r}
draw(p0x, p0y)
```

If all goes well, a simple sketch of Calvin should appear.

1.   

If you now draw the next set of vectors:

```{r}
sort.p1x <- sort(p1y)
sort.p1y <- sort(p1y)
draw(sort.p1x, sort.p1y) # returns a diagonal line
draw(p1x < 75, p2y < 75) # does not work
```

you will see Calvin and his mirror reflection, except that there is a problem with the data: some garbage points are mixed in with the correct data, resulting in a lot of spikes that reach reach between 75 and 100 in the y direction. Please fix this problem, so that drawing `p1x` with `p1y` results in a clean rendering of Calvin and his reflection.

1.   
```{r}
sort.p1y <- p1y[sort(p1y)]
sort.p1x <- p1x[sort(p1x)]
sort.p1y[sort.p1y > 75]
draw(sort.p1x[0:sort.p1x > 75], sort.p1y[0:sort.p2y > 75)
```

Let's take a look at the second example:

```{r add 4th Hobbes to upper right corner}
length(p2x)
length(p2y)

# reassign new variables to avoid re-loading data
new.p2x <- p2x
new.p2y - p2y

# take subset with missing hobbes and replace with columns containing 2
new.p2x[50:100]<- p2x[0:51]
new.p2y[50:100]<- p2y[0:51]
draw(new.p2y, new.p2x) # produces 2 hobbes re-oriented to face downward


# try switching x and y
new.p2x[50:100]<- p2y[0:51]
new.p2y[50:100]<- p2y[0:51]
draw(new.p2y, new.p2x) # produces 3 hobbes re-oriented to face downward

draw(p2x, p2y)
# instead of replace, try adding values
new.p2y<- p2x[0:50] +50
draw(new.p2x, p2y) # No go

new.p2x[50:100] <- new.p2x[]
```

here, we almost have four duplicates of Hobbes in a square configuration, but the upper right drawing is missing. Please duplicate one of the drawings of Hobbes, and place it in the upper right corner (by modifying the `p2x` and `p2y` vectors, that is).

1.   
```{r }
p2x[50:length(p2x)]<- p2x[0:50]
draw(new.p2x, p2y)
```

Let's take a look at the third example:

```{r}

draw(p3x, p3y)
```

here, Calvin is running around a circle, but the drawing closest to the origin is backwards. In particular, the x and y coordinates need to be flipped to restore symmetry. As before, please fix this problem.


<div id="license">
Adapted from material developed by Olaf Hall-Holt and Paul Roback for CS 125 at St. Olaf College.
</div>